---
title: "Analysis of Possible Demographic Factors Influencing Report of Mental Health Act Apprehensions in Toronto Area"
title-block-banner: true
author: 
  - Mingjia Chen
thanks: "Code and data are available at: https://github.com/MjChen120/Mental_Health_Apprehensions_Toronto."
date: today
date-format: long
abstract: "To have a general idea in terms of statistics of Mental Health Act (MHA) Apprehensions in the Toronto area, two of the datasets publically available from the Toronto Open Data Portal are speculated by using histograms and scatter plots. In general, the number of apprehensions increases by year and a decrease in the number starts in 2022, males in the age cohort 25-34 are more likely to have mental health-related apprehensions, and the total number of crimes and MHA apprehensions might also have indirect correlation within the same area. Although possible trends are detected, no relations are analyzed at this point. Even though various literatures have been published regarding mental health status, most of them lack a prespective in terms of law and crime: this report proposes to detect possible trends for future inspection and research regarding Mental health-related Apprehensions and possible contributing variables."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

#### Workspace set-up ####

## Load packages
library(AER)
library(future)
library(gitcreds)
library(knitr)
library(lintr)
library(renv)
library(reprex)
library(styler)
library(tictoc)
library(tidyverse)
library(tinytex)
library(usethis)
library(opendatatoronto)
library(janitor)

#### Read in cleaned data ####
cleaned_mentalHealth_data <- read.csv("cleaned_mentalHealth_data.csv")

cleaned_crime_data <- read.csv("cleaned_crime_data.csv")

summerized_report_data <- read.csv("summerized_report_data.csv")

```

## Introduction

By statistics, in any given year, 1 in 5 Canadians experiences a mental illness. By the time Canadians reach 40 years old of age, 1 in 2 have - or have had - a mental illness [@CAMH]. Mental health issues are not uncommon and should be treated seriously just like other types of health issues. Toronto is one of the major cities in Canada, which is also the capital of Province Ontario. Toronto fields includes old 140 City of Toronto Neighborhoods structure as well as the new 158 City of Toronto Neighborhoods structure. Large cities are usually associated with fast pace of life, inflation, competition, and high pressure; this large city draws our attention regarding the mental health status of its residents. What are the demographic factors of the population that is being apprehended under the Mental Health Act (MHA) in Toronto city? This report is interested in possible factors that contributes to the numbers of apprehensions under MHA. 

The MHA apprehensions and reports data set from Open Data Toronto Portal [@opendatatoronto] could be hinting the overall well-being of the residents in the area. This particular report thus examines the MHA Apprehensions data set [@torontoMHA] provided by Toronto Public Data by using R language [@citeR] to provide a overview of the statistics both demographic and year-related in regards of the mental health apprehensions. Number of crimes and violence are usually associated with high testosterone level [@testPaper]; men in young adulthood have significant positive association with impulsive and violent criminal behavior. Demographic factors such as age and gender could thus play a key role of Mental Health related apprehensions or crime reports in general. To elaborate, the relationships between number of reports and year, gender, and age are studied in this report. 

To hypothesize, the demographic factors gender and age correlates with the number of MHA Apprehensions. In addition, the number of Mental Health Apprehensions could be in some way related to total number of reported crimes in the same neighborhood. Thus, a supplementary data set called Police Annual Statistical Report [@torontoPoliceAnnual] regarding arrested and charged persons from Open Data Toronto [@opendatatoronto] is loaded for comparing the number of crimes reported in total and number of MHA apprehensions in the same area division to see if there is any correlation between the two. 

In result, potential trends in terms of relation between the demographic factors and crime rate in the same area division were detected. The relations were not statistically analyzed at this point of time, however, this report provides a foundation for research studies regarding the similar topics for Toronto area in the future. This report includes the introduction of the data sets in Data section, demographic analysis in Results section, and a Discussion section with limitations at the end.


## Data {#sec-data}

For the report, data used are retrieved from public online resource called Open Data Toronto Portal through the library `opendatatoronto` [@opendatatoronto].The data selected for the purpose of analysis are two package data sets called "Mental Health Act Apprehensions" [@torontoMHA] and "Police Annual Statistical Report - Arrested and Charged Persons" [@torontoPoliceAnnual]. Both of the data sets are measured mainly based on reported made from facilities. We downloaded, cleaned and tidied the data set using the statistical programming language R [@citeR] including the `tidyverse` [@tidyverse], `janitor` [@janitor], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr], `tibble` [@tibble], `kableExtra` [@kableExtra], and `knitr` [@knitr].

### Demographics of Mental Health Act Apprehensions

The primary data from "Mental Health Act Apprehensions" captures all Mental Health Act (MHA) Apprehensions pursuant to the MHA [@torontoMHA]. The data consists 25 variables, including the gender, age cohort, and year of the MHA apprehension reports. This particular data is chosen for the interest of the paper and no similar data sets are available from Open Data Toronto. Those three variables in particular and with the counting of the number of the cases will be used for demographic analysis. Each event number in the data set is a distinct MHA Apprehension and the data is queried based on report dates (@tbl-cleaned-mentalHealth-data).

```{r}
#| echo: false
#| message: false
#| label: tbl-cleaned-mentalHealth-data
#| tbl-cap: Sample of cleaned Mental Health Act Apprehensions Data

#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_mentalHealth_data, 5) |>
  kable(
    col.names = c("ID", "Report-Year", "Division", "Gender", "Age-Cohort"),
    booktabs = TRUE
  )

```

### Police Annual Statistical Report

The supplementary data set "Police Annual Statistical Report - Arrested and Charged Persons" is used for further inspection of the MHA apprehension data. The data set is chosen for the purpose of the study not only because there are no other similar data sets in the Open Data Toronto Portal, it is also updated by the same publisher as the MHA apprehension data set; thus, the format is consistent with the primary data. The data provides the aggregate count of people who have been arrested and charged by police station division, neighborhood, sex (or gender), age, crime category, and crime sub-type [@torontoPoliceAnnual]. The data set contains 11 variables in total. The variables unique data id, year of arrest, division of police station, and number of arrests are thus included for this purpose (@tbl-cleaned-crime-data). 

```{r}
#| echo: false
#| message: false
#| label: tbl-cleaned-crime-data
#| tbl-cap: Sample of cleaned Police Annual Statistical Report Data

#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
head(cleaned_crime_data, 5) |>
  kable(
    col.names = c("ID", "Arrest-Year", "Division", "Arrest-Count"),
    booktabs = TRUE
  )
```

### Combining Datasets

By assumption, areas with more crime reports should have more mental health related apprehensions due to the fact that MHA apprehensions are part of the crime reports; however, mediators and moderators might take place to impact the relationship between these two variables. For instance, residents in an area that has high crime rate could be violating the laws due to mental pressures, which could potentially trigger mental health issues. Crimes around an individual might negatively affect their mental health, thus influencing them into committing mental health apprehensions. As explained, this particular data package is thus included in this report for finding possible factors that affect the mental health apprehension in one area.

The Annual Police Report data [@torontoPoliceAnnual] is thus combined with the MHA Apprehension data [@torontoMHA] to compare the number of apprehensions under MHA and number of arrests within the same year and same police station division to study a possible pattern. @tbl-summerized-data presents a snapshot of the finalized combined data.

```{r}
#| echo: false
#| message: false
#| label: tbl-summerized-data
#| tbl-cap: Sample of Summerized data combining the above two datasets

#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd
demo <- summerized_report_data |> group_by(year, division)

head(demo,10) |>
  relocate(any_of(c("X","year","division","counts","type"))) |>
    kable(
      booktabs = TRUE,
      col.names = c("ID", "Arrest-Year", "Division", "Counts", "Type"),
      align = c("c", "c", "c", "c", "c")
        )
```

## Results

### Number of Reports by Years

The year of the reports and numbers of reports recorded in a year is being visualized by a histogram. The data is cleaned again for this purpose specifically to remove parts of the data that lack gender and age cohort reports. As we can see, the number increases as the year increases until 2022. After 2022, the number of apprehensions decreases (@fig-years). This might be an indicator where residents' mental healthy generally becomes better after 2022. The differences between year categories might not necessarily be significant at this point of analysis. In comparison, the total of number of arrests in general decreases through the years with a increase until the year of 2020. Although the trend of the data both changed, no scientific evidence supports the theory that the number of reports might be influenced by Covid-19 pandemic; however, the difference in the trends of the data over the years might hints the insight that the number of MHA apprehensions might not be directly correlated to the number of total arrests.

```{r}
#| label: fig-years
#| fig-cap: Counts of Reports in Each Year
#| fig-subcap: ["MHA Reports","Arrest Reports"]
#| layout-ncol: 2
#| echo: false

cleaned_mentalHealth_data|>
  ggplot(aes(x = year)) +
  geom_bar() +
  theme_minimal() + # Make the theme neater
  labs(x = "Year", y = "Number of reports") 

report_data <- summerized_report_data |> filter(type == "ALL") |> group_by(year) |> summarise_at(vars(counts), list(name=sum)) |> ungroup()
names(report_data )[names(report_data ) == "name"] <- "counts"
ggplot(report_data, aes(x = year, y = counts)) +
       geom_bar(stat="identity") + 
  theme_minimal() + # Make the theme neater
  labs(x = "Year", y = "Number of reports") 

```

\break

### Demographic Factors

```{r,include=FALSE}
#clean unnecessary NA reports only for demographic analysis

data <- cleaned_mentalHealth_data[!cleaned_mentalHealth_data$sex == "Not Recorded",]
data <- data[!data$age_cohort == "Not Recorded",]

male_per <- data[data$sex == "Male",]
male_per <- round(100 * nrow(male_per) / nrow(data), digits = 2) 

age_per <- data[data$age_cohort == "25 to 34",]
age_per <- round(100 * nrow(age_per) / nrow(data), digits = 2)

```

With sketching histograms, the number of reports in different groups of population are compared to each other based on gender and age. In addition, the four histograms are drawn to visualize the difference between those groups (@fig-groups and @fig-combined). As result, males with age from 25 to 34 are more likely to report mental health apprehensions (`r male_per`% males, `r age_per`% with age cohort from 25 to 34). The differences between gender and age group are not tested in terms of significance even though it could be significant with high power level due to large sample size. The analysis has limitations such as it only includes binary gender (i.e., female vs. male); further limitations will be discussed after in the discussion section.

```{r}
#| label: fig-groups
#| fig-cap: Gender and Age of People Apprehended
#| fig-subcap: ["Gender","Age"]
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false

data |>
  ggplot(aes(x = sex)) +
  geom_histogram(stat = "count")+
  theme_minimal() + # Make the theme neater
  labs(x = "Gender", y = "Number of reports") 

data |>
  ggplot(aes(x = age_cohort)) +
  geom_histogram(stat = "count")+
  theme_minimal() + # Make the theme neater
  labs(x = "Age Cohort", y = "Number of reports") 

```

```{r}
#| label: fig-combined
#| fig-cap: Gender and Age of People Apprehended Visulization Year-Depended
#| fig-subcap: ["Gender","Age"]
#| layout-ncol: 2
#| echo: false
#| warning: false
#| message: false


data |>
  ggplot(aes(x = year, fill = sex)) +
  geom_histogram(position = "dodge") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Year",
    y = "Occurrences",
    fill = "Sex"
  )

data |>
  ggplot(aes(x = year, fill = age_cohort)) +
  geom_histogram(position = "dodge") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") +
  labs(
    x = "Year",
    y = "Occurrences",
    fill = "Age Cohort"
  )

```
\break

### Number of MHA versus Total Reports per Division

To visualize the number of reports and arrests in an area, average counts of MHA apprehensions and Police Reports over the years for each division is compared against each other. Again, as mentioned, MHA reports will always be less than total reports since it is a part of the reports at the first place; however, as demonstrated in @fig-merged, the percentage of MHA reports among total reports are different from each other. For instance, larger numbers of police reports are intuitively associated with larger numbers of MHA reports (i.e., Division 13 versus Division 14). Results, however, shows that in some cases where one division has larger numbers of police reports but smaller numbers of MHA reports compared to others (i.e., Division 14 versus Division 51). The percentages of average MHA reports out of average police reports of the years are not consistent across the divisions as well. The report does not test the significance of the difference and lacks enough information for detecting possible moderators and mediators, thus, we can only state that there could be an indirect correlation between the number of total crimes and MHA reports within the same division in the same area.  

```{r}
#| label: fig-merged
#| fig-cap: Gender and Age of People Apprehended
#| echo: false
#| warning: false
#| message: false

# modify data only for graphing purpose 
data_mha <- summerized_report_data[summerized_report_data$type == "MHA",] |> group_by(division) |> summarise_at(vars(counts), list(counts=sum)) |> ungroup()
data_mha$type = rep("MHA", times = nrow(data_mha))

data_all <- summerized_report_data[summerized_report_data$type == "ALL",] |> group_by(division) |> summarise_at(vars(counts), list(counts=sum)) |> ungroup()
data_all$type = rep("ALL", times = nrow(data_all))

merged_data <- rbind(data_mha,data_all)

#graphing data
merged_data|>
  ggplot(aes(x = division, y = counts, color = type)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Division", y = "Number of Reports and Arrests", color = "Type") +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom")

```

\newpage

## Discussion and Conclusion

Overall, the data reveals relationships between demographic factors (age, gender, and area or neighborhood) and number of MHA apprehensions reports. By conducting histogram graphs, We found that the number of MHA reports increases each year until 2022: in comparison, the total crime reports in Toronto decreases until 2020 (@fig-years). This trend of data could be a result of Covid-19 Pandemic due to the fact that many people encounter mental health issues while staying mostly at homes. The difference of the patterns in MHA and total crime reports also hints a potential indirect correlation with each other. Demographically, males in age cohort from 25 to 34 are more likely to be reported as MHA apprehensions(`r male_per`% males, `r age_per`% with age cohort from 25 to 34) as shown in @fig-groups. The difference between gender and age groups are not necessarily significant though are most likely to be since the large sample size. Closer studies such as model and power analysis should be conducted for better statistical understanding of the data sets.

Additionally, an possible indirect relation between the number of MHA apprehension and corresponding area division is detected: increase in number of total crime in one division does not correlates to higher number of MHA apprehensions with the same degree within areas (@fig-merged). The percentages of average MHA reports out of average police reports of the years are not consistent across the divisions. In other words, possible bias, mediating, or moderating factors could take place in the correlation equation such as @eq-relation. Although the paper does not create and testify the validity of the equations and models, it provides a potential future direction for the research field.

$$
Number Of MHA = r * (Total Number Of Crimes) + Other factors + Bias
$$ {#eq-relation}

### Weaknesses and Next Steps {#sec-first-point}

There are many weaknesses and limitation of the research. For MHA apprehension data, a unique event recorded in the data does not represent as an apprehension of a distinct individual: an individual may have been apprehended multiple times under the Mental Health Act, with those apprehensions recorded each with a unique event number in the data set. This fact probably applies to the Police Annual report data set as well, even though not mentioned in its description. This fact indicates that the visualization in the report does not represent number of people who are apprehended or arrested, but rather how many times those events happened. That could potentially affect the data and how to interpret data as one individual is responsible for a large amount of the apprehensions and crimes. For instance, the reason why males from 25 to 34 years old are generally apprehended more in the data (@fig-groups) could be due to one male being apprehended multiple times. Furthermore, MHA Apprehensions data recorded events from year 2014 to 2022, whereas the crime data set only has events from year 2014 to 2023. The lack of one year data could dramatically change the outlook of the data visualizations due to the fact that only averaged number of reports over the years per division from MHA and crime data sets were used (@fig-years).

Other than limitations, the exact relationships are not evaluated statistically in this report; thus, in terms of future directions, actual correlations and relations can be analyzed by conducting proper statistical analysis such as Power or Model analysis. Better way of tidying and cleaning data could be done for current data limitations. Additional data sets and field studies could as well be conducted and used for supplementing the statistical analysis regarding the topic. 
\newpage

\appendix

## Appendix {-}

### Additional data details
This appendix includes the snapshots of the data sets cleaned, tidied, and used for the purpose of the research. 
The script for data cleaning and the actual raw and cleaned data sets are available at https://github.com/MjChen120/Mental_Health_Apprehensions_Toronto

#### Overview of Data Mental Health Apprehsions 

```{r}
#| echo: false
#| message: false
#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd

head(cleaned_mentalHealth_data,15) |>
  relocate(any_of(c("X","year","division","sex","age_cohort"))) |>
    kable(
      booktabs = TRUE,
      col.names = c("ID", "Report-Year", "Division", "Gender", "Age cohort"),
      align = c("c", "c", "c", "c", "c")
        )
```

#### Overview of Data Police Annual Report

```{r}
#| echo: false
#| message: false
#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd

head(cleaned_crime_data,15) |>
  relocate(any_of(c("X","year","division","counts"))) |>
    kable(
      booktabs = TRUE,
      col.names = c("ID", "Arrest-Year", "Division", "Counts"),
      align = c("c", "c", "c", "c", "c")
        )
```

#### Overview of Data Sets Summerized and Combined 

The additional data set used for analysis is combined by R. All the number of MHA reports in the year for the same division is counted as the variable "counts" in MHA Apprehension data set; whereas all the reported numbers of the crimes for same division in the same year are summed as a column of "counts" as well in Police annual reports. The two data sets are then combined together with an extra column indicating the type of the reports they are from to differentiate (for example, "ALL" indicates the counts is for general police report, and "MHA" indicates the counts is for MHA report. 

```{r}
#| echo: false
#| message: false
#referenced kable stying from https://github.com/EthanSansom/torontocitytrees/blob/main/outputs/paper/paper.Rmd

head(summerized_report_data,15) |>
  relocate(any_of(c("X","year","division","counts","type"))) |>
    kable(
      booktabs = TRUE,
      col.names = c("ID", "Arrest-Year", "Division", "Counts", "Type"),
      align = c("c", "c", "c", "c", "c")
        )
```

\newpage
## References
